---
title: "Vignette2"
author: "Thomas Goralski"
date: "2025-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```









The commented out lines detail how the NMF embeddings were computed. This takes a large amount of compute, so for efficiency the objects have been saved after nmf embedding and can be loaded directly in.

# ```{r load geomx data and create suerat object}
# 
# #define file paths
# file_paths <- list(
#   "ACA_MOS_MOp_6mpi" = "/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/target_data_ACA_MOS_MOp_6mpi.RDS",
#   "ACA_MOS_MOp_9mpi" = "/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/target_data_ACA_MOS_MOp_9mpi.RDS",
#   "LAMDA" = "/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/target_data_LAMDA.RDS",
#   "MultiRegion" = "/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/target_data_MultiRegion.RDS",
#   "SNcAmygNew" = "/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/target_data_SNcAmygNew.RDS"
# )
# 
# 
# 
# #make suerat object
# seurat_object <- combine_datasets_into_seurat(file_paths)
# 
# 
# ```
# 
# 
# 
# ```{r get nmf embeddings}
# seurat_object <- process_and_run_nmf(seurat_object, nmf_rank = 5)
# RankPlot(seurat_object, detail.level = 2)
# 
# ```
# 
# 
# ```{r load reference Allen data}
# 
# Allen_SEA_AD_Atlas <- readRDS("/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/Reference_MTG_RNAseq_all-nuclei.2022-06-07.rds")
# 
# ```
# 
# 
# ```{r run NMF on allen data}
# Allen_SEA_AD_Atlas <- process_and_run_nmf(Allen_SEA_AD_Atlas, nmf_rank = 5)
# 
# ```
# 
# 
# ```{r save objects}
# saveRDS(seurat_object,"seurat_object.rds")
# saveRDS(Allen_SEA_AD_Atlas,"Allen_SEA_AD_Atlas.rds")
# ```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


The commented out lines detail how the NMF embeddings were computed. This takes a large amount of compute, so for efficiency the objects have been saved after nmf embedding and can be loaded directly in.

```{r load geomx data and create suerat object, eval=FALSE}

#define file paths
file_paths <- list(
  "ACA_MOS_MOp_6mpi" = "/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/target_data_ACA_MOS_MOp_6mpi.RDS",
  "ACA_MOS_MOp_9mpi" = "/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/target_data_ACA_MOS_MOp_9mpi.RDS",
  "LAMDA" = "/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/target_data_LAMDA.RDS",
  "MultiRegion" = "/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/target_data_MultiRegion.RDS",
  "SNcAmygNew" = "/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/target_data_SNcAmygNew.RDS"
)



# Create a Seurat object by combining multiple datasets
seurat_object <- combine_datasets_into_seurat(file_paths)


```

```{r get nmf embeddings, eval=FALSE}
# NMF (Non-negative Matrix Factorization) is used for dimensionality reduction
# Setting the rank to 5 means choosing to reduce the data dimensions into 5 factors/components, which will each capture a distinct pattern within the data to explain variability in the original dataset
seurat_object <- process_and_run_nmf(seurat_object, nmf_rank = 5)

# Plot the embeddings of NMF rank 2 
RankPlot(seurat_object, detail.level = 2)
```


```{r load reference Allen data, eval=FALSE}

Allen_SEA_AD_Atlas <- readRDS("/varidata/research/projects/henderson/THOMAS/Machine_Learning_For_Bioinformatics_Project/Reference_MTG_RNAseq_all-nuclei.2022-06-07.rds")

```
 
 
 
```{r run NMF on allen data, eval=FALSE}
Allen_SEA_AD_Atlas <- process_and_run_nmf(Allen_SEA_AD_Atlas, nmf_rank = 5)
```


```{r save objects, eval=FALSE}
saveRDS(seurat_object,"seurat_object.rds")
saveRDS(Allen_SEA_AD_Atlas,"Allen_SEA_AD_Atlas.rds")
```


how the subset object was made
```{r load in nmf embedded data, eval=FALSE}

 seurat_object <- readRDS("/varidata/research/projects/henderson/THOMAS/SpatialIntegration/SpatialIntegration/seurat_object.rds")
Allen_SEA_AD_Atlas <- readRDS("/varidata/research/projects/henderson/THOMAS/SpatialIntegration/SpatialIntegration/Allen_SEA_AD_Atlas.rds")

# Replace NA values in the "subclass_label" column of @meta.data with "NA"
Allen_SEA_AD_Atlas@meta.data[["subclass_label"]][is.na(Allen_SEA_AD_Atlas@meta.data[["subclass_label"]])] <- "NotLabled"

```

```{r subset seurat object for upload to data file in package, eval=FALSE}
#Subset the data by proportions to upload a smaller datafile to the package
seurat_object_sub<-subset_seurat_by_proportions(seurat_object = seurat_object, cell_type_col="segment", 500)
allen_sub<-subset_seurat_by_proportions(seurat_object = Allen_SEA_AD_Atlas, cell_type_col="subclass_label", 6000)

```

```{r, eval=FALSE}
 # merge the subset data for allen and seurat into one dataset
  seurat_merge<-merge_seurat_objects(seurat_object = seurat_object_sub, allen_sub)
 
  #save data to data file
  use_data(seurat_merge, compress = "bzip2", overwrite = TRUE)
 
```


