---
title: "Spatial_Integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial_Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SpatialIntegration)
```



Parkinson’s disease (PD) is a neurodegenerative disease characterized by intraneuronal ɑ-synuclein aggregates known as Lewy pathology. Using GeoMx bulk spatial transcriptomics, we identified a molecular signature, Lewy-Associated Molecular Dysfunction from Aggregates (LAMDA), consisting of downregulated DNA repair, mitochondrial, and synaptic pathways in cortex of post-mortem PD and mouse model brains. Previous single-nucleus (sn) transcriptomics studies in PD lacked resolution on aggregate status, making it unclear how sn alterations align with the LAMDA signature. This package enables integration of bulk spatial and sn transcriptomics data and apply non-negative matrix factorization to learn shared features across data types in a low-dimensional space (aim 1) and transfer “spatial” labels from the bulk to sn data by machine learning (aim 2).


```{r}

#load the data
data(seurat_merge)

```

```{r}
is(seurat_merge, "Seurat")

seurat_merge
```


Find contribution to overall dataset to avoid NMF fitting to only 1 data set
```{r}
#get indiceds of geomx data
geo_ind<-is.na(seurat_merge@meta.data[["specimen_name"]])

#get sum of geomx data
sum(seurat_merge@assays[["RNA"]]@layers[["counts"]][geo_ind])
#get sum of allen data
sum(seurat_merge@assays[["RNA"]]@layers[["counts"]][-geo_ind])

```

The process and run nmf function will do the rescaling for us (allen is shrunk to scale to geomx)


```{r}

#for whatever reason, the singlet package will not load as a dependency, so load it here
library(singlet)
```

```{r}

seurat_merge<-process_and_run_nmf(seurat_merge, nmf_rank = 15)

```


```{r}


seurat_merge<-perform_gsea_and_clustering(seurat_merge)


```


```{r}
#View umap clusters
print(DimPlot(seurat_merge, reduction = "umap"))
#View GeoMx segment lables
print(DimPlot(seurat_merge, reduction = "umap", group.by ="segment"))
#View Cell subclass labels from allen
print(DimPlot(seurat_merge, reduction = "umap", group.by ="subclass_label"))

#Veiw Cells labled NA in original Allen data
print(DimPlot(seurat_merge, reduction = "umap", group.by ="class_label"))
```


```{r}
#View each cell type label and spatial annotations correlation to each nmf factor
plot_nmf_correlation_heatmap(seurat_merge, "cell_type_nmfFactor_corr.png")
```

```{r}
#View nmf loadings on umaps
FeaturePlot(seurat_merge, features = paste0("NMF_", c(2,3,7,8,c(12:15))), raster = FALSE)

```



```{r}

#get genesets that load with NeuN related factor
plot_top20_enriched_pathways_single(seurat_object = seurat_merge,
  nmf_factor = 13,
  output_file = "top20_enriched_nmf6.pdf"
)

#get genesets that load with pSyn related factor
plot_top20_enriched_pathways_single(seurat_object = seurat_merge,
  nmf_factor = 2,
  output_file = "top20_enriched_nmf15.pdf"
)


```



```{r plot GSEA and NMF features}

GSEAHeatmap(seurat_merge, reduction = "nmf", max.terms.per.factor = 50)
FeaturePlot(seurat_merge, features = paste0("NMF_", 1:10), raster = FALSE)
DimPlot(seurat_merge, reduction = "umap", group.by ="segment")
```



```{r retrieve references}
# Install RefManageR if not already installed
if (!requireNamespace("RefManageR", quietly = TRUE)) {
  install.packages("RefManageR")
}

# Load the package
library(RefManageR)

# Set the path to your .bib file (adjust if you moved it)
bib_file <- "/varidata/research/projects/henderson/THOMAS/SpatialIntegration/SpatialIntegration/references.bib"  
file.exists(bib_file)

# Read the BibTeX file
bib_entries <- ReadBib(bib_file)

# Print the loaded references
print(bib_entries)

```


[1] D. Agarwal, C. Sandor, V. Volpato, et al. “Common gene expression
signatures in Parkinson’s disease are driven by changes in cell composition”.
In: _npj Parkinson's Disease_ (2020). DOI: 10.1038/s41531-020-0112-5.
<https://pubmed.ncbi.nlm.nih.gov/32317022/>.

[2] R. Feleke, R. Reynolds, A. Smith, et al. “Cross-platform transcriptional
profiling identifies common and distinct molecular pathologies in Lewy body
diseases”. In: _Acta Neuropathologica_ 142 (2021), pp. 449-474. DOI:
10.1007/s00401-021-02343-x. <https://doi.org/10.1007/s00401-021-02343-x>.

[3] H. Fernandes, N. Patikas, S. Foskolou, et al. “Single-Cell Transcriptomics
of Parkinson’s Disease Human In Vitro Models Reveals Dopamine Neuron-Specific
Stress Responses”. In: _Cell Reports_ 33 (2020), p. 108263. DOI:
10.1016/j.celrep.2020.108263. <https://doi.org/10.1016/j.celrep.2020.108263>.

[4] T. Goralski, L. Meyerdirk, L. Breton, et al. “Spatial transcriptomics
reveals molecular dysfunction associated with cortical Lewy pathology”. In:
_Nature Communications_ 15 (2024), p. 2642. DOI: 10.1038/s41467-024-47027-8.
<https://doi.org/10.1038/s41467-024-47027-8>.

[5] M. van den Hurk, S. Lau, M. Marchetto, et al. “Druggable transcriptomic
pathways revealed in Parkinson’s patient-derived midbrain neurons”. In: _npj
Parkinson's Disease_ 8 (2022), pp. 1-18. DOI: 10.1038/s41531-022-00400-0.
<https://doi.org/10.1038/s41531-022-00400-0>.

[6] L. Yang, H. Qian, Y. Qin, et al. “scGPT: toward building a foundation model
for single-cell multi-omics using generative AI”. In: _Nature Biotechnology_
(2024). Accessed via PubMed. DOI: 10.1038/s41587-024-02238-2.
<https://pubmed.ncbi.nlm.nih.gov/38409223/>.





